---
title: "Interactive graphics and animation"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.height=3,
                      fig.align='center',
                      tidy.opts=list(width.cutoff=60),
                      tidy=FALSE, 
                      warning = FALSE,
                      message = FALSE)
library(dplyr)
library(ggplot2)
library(plotly)
library(tidyr)
library(sf)
library(gapminder)
library(rnaturalearth)
library(gganimate)
library(knitr)
```

```{r wrap-hook, echo=FALSE}
# text wrapping for nice slides
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = knitr:::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})
knitr::opts_chunk$set(linewidth=60)
```

# Motivating example

## Matric pass rates

```{r, echo = FALSE, fig.height=5}
matrics <- data.frame(
  year = 2014:2018,
  EC = c(65.4,56.8,59.3,65,70.6),
  FS = c(82.8,81.6,88.2,86.1,87.5),
  Gau = c(84.7,84.2,85.1,85.1,87.9),
  KZN = c(69.7,60.7,66.4,72.9,76.2),
  Lim = c(72.9,65.9,62.5,65.6,69.4),
  Mpu = c(79,78.6,77.1,74.8,79),
  NW = c(84.6,81.5,82.5,79.4,81.1),
  NC = c(76.4,69.4,78.7,75.6,73.3),
  WC = c(82.2,87.7,85.9,82.8,81.5)
)

matrics_long <- pivot_longer(matrics, cols = -1, names_to = "province", values_to = "pass_rate")

# ggplot2 
p <- matrics_long %>% 
  ggplot(aes(x = year, y = pass_rate, colour = province)) +
  geom_line() + geom_point() + 
  scale_colour_brewer(palette = "Set3")
p
```

## Making it interactive

```{r, echo = FALSE, fig.height=5}
p <- ggplotly(p)
p
```

## Animating it

```{r, echo=FALSE, out.width=700}
ggplot(matrics_long, aes(year, pass_rate, group = province)) +
  geom_line() +
  geom_segment(aes(xend = 2018, yend = pass_rate), linetype = 2, colour = 'grey') +
  geom_point(size = 2) +
  geom_text(aes(x = 2018.1, label = province), hjust = 0) +
  transition_reveal(year) +
  coord_cartesian(clip = 'off') +
  labs(title = 'Matric pass rates in SA', y = 'Pass rate (%)') +
  theme_minimal()
```

# How does it work?

## A slightly easier example

```{r, tidy = FALSE}
p <- diamonds %>% sample_n(1000) %>%
  ggplot(aes(x = x, y = price)) + geom_point() +
  labs(x = "Diamond width (x-dim)", y = "Price")
p
```

## Make interactive with `plotly`

`plotly` tries to guess the right plot type

```{r, tidy = FALSE}
diamonds %>% sample_n(1000) %>%
  plot_ly(x = ~x, y = ~price)
```

## Make interactive with `plotly`

Better to specify with `add_[plottype]`

```{r, tidy = FALSE}
diamonds %>% sample_n(1000) %>%
  plot_ly(x = ~x, y = ~price) %>%
  add_markers()
```

## Make interactive with `plotly`

Styling options use `plotly` syntax

```{r, tidy = FALSE}
diamonds %>% sample_n(1000) %>%
  plot_ly(x = ~x, y = ~price) %>%
  add_markers(color = I("red")) %>%
  layout(xaxis = list(title = "Diamond width (x-dim)"),
         yaxis = list(title = "Price"))

```

## Make interactive with `plotly`

"Aesthetic" options also use `plotly` syntax

```{r, tidy = FALSE}
diamonds %>% sample_n(1000) %>%
  plot_ly(x = ~x, y = ~price, color = ~clarity, colors = "Set2") %>%
  add_markers() %>%
  layout(xaxis = list(title = "Diamond width (x-dim)"),
         yaxis = list(title = "Price"))
```

## Exercise

Open `01-ia-exercise.R`.

Make the plot below interactive using `plotly`. Reproduce as much of the plot as you can (colours, titles, etc)

```{r, echo=FALSE}
diamonds %>% count(cut) %>%
  ggplot(aes(x=cut, y=n)) + geom_bar(stat = "identity", fill = "red") +
  ggtitle("Number of diamonds per cut category")
```

## The easy way with `ggplotly`

Step 1: Make the `ggplot`

```{r, eval = FALSE, tidy = FALSE}
# with colour by clarity
p <- diamonds %>% sample_n(1000) %>%
  ggplot(aes(x = x, y = price, colour = clarity)) +
  geom_point() + geom_smooth(se = FALSE) +
  labs(x = "Diamond width (x-dim)", y = "Price") +
  scale_color_brewer(palette = "Set2")
p
```

## The easy way with `ggplotly`

Step 1: Make the `ggplot`

```{r, echo = FALSE}
# with colour by clarity
p <- diamonds %>% sample_n(1000) %>%
  ggplot(aes(x = x, y = price, colour = clarity)) +
  geom_point() + geom_smooth(se = FALSE) +
  labs(x = "Diamond width (x-dim)", y = "Price") +
  scale_color_brewer(palette = "Set2")
p
```

## The easy way with `ggplotly`

Step 2: `ggplotly` the ggplot!

```{r}
ggplotly(p)
```

## Exercise

Open `02-ia-exercise.R`.

Differences between clarity groups in the previous plot might be easier to see if we plot *densities* rather than all points. Replot the previous plot as a 2-D density plot, coloured by clarity, using the `geom_density2d` geom.

## Interactive maps

```{r, tidy=TRUE}
# one of the plots we made in the previous lecture
occupancies <- st_read("../spatial-data/data/Mongolia_SL.shp", quiet = TRUE)
sights <- read.csv("../spatial-data/data/confirmed-sightings.csv")
sights <- st_as_sf(sights, coords = c("Longitude", "Latitude"), crs = 4326) %>%
  st_transform(sights, crs = "+proj=utm +zone=48 +datum=WGS84 +units=m +no_defs")

p <- ggplot() + geom_sf(data = occupancies, aes(fill = Occ)) +
  geom_sf(data = sights, aes(colour = Elevation)) +
  scale_colour_distiller(palette = "PRGn") +
  scale_fill_distiller(palette = "YlOrRd")
```

## Interactive maps

```{r, echo = FALSE, fig.height=5}
p
```

## Interactive maps

```{r, echo = FALSE, fig.height=5}
ggplotly(p)
```

# Animation

## How does it work?

Uses "frame" argument to index time
```{r}
df <- data.frame(x = runif(1000), y = runif(1000), f = 1:1000)
p <- ggplot(df, aes(x, y)) + geom_point(aes(frame = f))
p
```

## How does it work?

No frame argument for `ggplot` but `ggplotly` knows what to do
```{r}
ggplotly(p)
```

## Back to matric pass rates

```{r}
p <- matrics_long %>%
  ggplot(aes(x = year, y = pass_rate, colour = province)) +
  geom_line(aes(frame = year)) + geom_point(aes(frame = year)) +
  scale_colour_brewer(palette = "Set3")
p
```

## Back to matric pass rates

```{r}
ggplotly(p)
```

## More advanced example

```{r}
pgap <- ggplot(
  gapminder,
  aes(x = gdpPercap, y=lifeExp, size = pop, colour = country)) +
  geom_point(aes(frame = year), show.legend = FALSE, alpha = 0.7) +
  scale_color_viridis_d() +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  labs(x = "GDP per capita", y = "Life expectancy")
```

## More advanced example

```{r}
pgap
```

## More advanced example

```{r}
ggplotly(pgap)
```

## Or with `gganimate` package

```{r, out.width=700}
pgap + transition_time(year) +
  labs(title = "Year: {frame_time}")
```

## How does it work?

Adds **transitions** - functions that determine how the plot data should be distributed over frames

+ `transition_time`
+ `transition_states`
+ `transition_reveal`
+ `transition_manual`
+ & others

Uses **tweening** - calculates intermediate data for a smooth transition between the states

**Easing** controls how these intermediate data are calculated (defaults to linear)

## `transition_states`

Transition between several *distinct stages* of the data

+ `states`: column with state levels.
+ `transition_length`: length of the transition between states.
+`state_length`:	length of the pause at the states.
+ `wrap = TRUE`: if TRUE the last state will be transitioned into the first.

## `transition_states`

```{r, out.width=700}
pgap <- ggplot(
  gapminder,
  aes(x = gdpPercap, y=lifeExp, size = pop, colour = year)) +
  geom_point(show.legend = FALSE, alpha = 0.7) +
  scale_color_viridis_c() +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  labs(x = "GDP per capita", y = "Life expectancy") +
  transition_states(continent) +
  labs(title = "Continent: {closest_state}")
```

## `transition_states`

Something not quite right
```{r, echo=FALSE, out.width=700}
pgap
```

## `transition_states`

Need to add `group = continent` to `aes()`
```{r, out.width=700}
pgap2 <- ggplot(
  gapminder,
  aes(x = gdpPercap, y=lifeExp, size = pop, colour = year, group = continent)) +
  geom_point(show.legend = FALSE, alpha = 0.7) +
  scale_color_viridis_c() +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  labs(x = "GDP per capita", y = "Life expectancy") 
```

## `transition_states`

```{r, echo=FALSE, out.width=700}
pgap2 + transition_states(continent, 
                          transition_length = 3,
                          state_length = 3) +
  labs(title = "Continent: {closest_state}") 
```

## `transition_reveal`

Reveal data `along` a given dimension
```{r, out.width=700}
p <- ggplot(matrics_long, aes(year, pass_rate, group = province)) +
  geom_line() +
  geom_point(size = 2)
p + transition_reveal(along = year)
```

## `transition_manual`

Maps a variable to a specific frame in the animation, with no tweening of data.

```{r, out.width=700}
pgap + transition_manual(year) +
  labs(title = "Year: {current_frame}")
```

## `transition_manual`

```{r, out.width=700}
pgap + transition_manual(year) +
  labs(title = "Year: {current_frame}") +
  facet_grid(~continent)
```

## Output and save animations

```{r, out.width=700}
animate(pgap, renderer = gifski_renderer(loop = TRUE))
```

```{r}
# defaults to last animation if `animation` option not set
anim_save(filename = "my_gap_gif.gif", animation = pgap)
```

## `gganimate` control settings

+ `transition_*()`: how data should change across time.
+ `view_*()`: how scales should change along the animation.
+ `shadow_*()`: how data from other points in time should be presented.
+ `enter_*()/exit_*()`: how new data should appear and how old data should disappear.
+ `ease_aes()`: how aesthetics should be eased during transitions.

## Exercise

Exercise `03-ia-exercise.R` takes you through various control settings for how `gganimate` plots transitions.

Exercise `04-ia-exercise.R` illustrates a spatial version of the gapminder animations we made earlier, so that we can plot animations of e.g. changes in life expectancy on top of a map.

## Additional resources

+ https://plot.ly/ggplot2/#animations
+ https://plotly-r.com/index.html
+ https://github.com/thomasp85/gganimate
+ https://cran.r-project.org/web/packages/gganimate/vignettes/gganimate.html
+ https://d4tagirl.com/2017/05/how-to-plot-animated-maps-with-gganimate
