---
title: "Exploring temporal data"
output: ioslides_presentation
---

```{r, include = FALSE}
knitr::opts_chunk$set(fig.height=3.5,
                      fig.align='center',
                      tidy.opts=list(width.cutoff=60),
                      tidy=TRUE, 
                      warning = FALSE,
                      message=FALSE)
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)
library(tsibble)
library(feasts)
# library(remotes)
# remotes::install_github("njtierney/brolgar")
library(brolgar) # not on CRAN
library(dygraphs)
library(knitr)
```

```{r wrap-hook, echo=FALSE}
# text wrapping for nice slides
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = knitr:::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})
knitr::opts_chunk$set(linewidth=60)
```

# Introducing dates, times <br /> and time series 

## Date and time types and formats

- Dates and times can be represented in different ways
<br><br>
- There are lots of different date and time formats 
<br><br>
- A date-time is a data type that includes a date and a time
<br><br>
- It can be frustrating getting your times into the right format
<br><br>
- I'll use R packages `lubridate` and `base`

## Time series data

- Events occurring sequentially through time can be thought of as time series
<br><br>
- These could be weather measurements, purchases, football match outcomes, even locations in space

## In this lecture  

- We will first go through different types of time and date classes
<br><br>
- Then we will do some visualisation and exploration of time series

## Dates, times and date-times

### Battle of the libraries: `base` vs `lubridate` 

- in `base` R dates and date-times have a special class called `posixCT`
  
```{r}
# date-time including time zone
Sys.time()
```

```{r}
# date in current time zone
Sys.Date()
```

## Dates, times and date-times
### Battle of the libraries: `base` vs `lubridate` 

- `lubridate` is a tidyverse package
- You can do what we did above using lubridate

```{r}
# date-time including time zone
now()       # lubridate
```

```{r}
# date in current time zone
today()     # lubridate
```

## Reading in date information

- This can be harder than it should be, especially when using tab- or comma-delimited formats

```{r}
wind_data <- read.csv("data/wind_direction.csv")
head(wind_data, n=3)
class(wind_data$Date)
```

## Changing the date format with `base`

- As you can see, R thinks the date column is a factor
- In `base` R you often need to specify the "Unix Epoch" <br /> 1970-01-01 in the origin argument

```{r, tidy = FALSE}
my_base_date <- as.Date(as.character(wind_data$Date), 
                        tryFormats = c("%d-%m-%Y", "%d/%m/%Y"), 
                        origin="1970-01-01")
head(my_base_date, n=4) 
class(my_base_date)
```

## Changing the date format with `lubridate`

- `lubridate` syntax is intuitive and compact

```{r}
class(wind_data$Date)

my_lbdt_date <- dmy(as.character(wind_data$Date))
head(my_lbdt_date, n=4) 
class(my_lbdt_date)
```

## Pitfalls

- The `as.character` bit is important!

```{r, tidy = FALSE}
as.Date(wind_data$Date, 
        format="%Y-%m-%d", tz="SAST", 
        origin="1970-01-01")

as.Date(as.character(wind_data$Date), 
        format="%d/%m/%Y", tz="SAST", 
        origin="1970-01-01")
```

## Pitfalls

- If you tell R the object contains times it will make them up

```{r}
my_fake_dt <- as.POSIXct(my_lbdt_date, format= "%Y-%m-%d", 
                                       tz="SAST", 
                                       origin="1970-01-01 00:00:00")
head(my_fake_dt, n=4)
```

## Pitfalls

- You can't add a timezone to a date
- Even if you do add it, like we did, it won't show up

```{r, tidy=FALSE}
my_base_date <- as.Date(as.character(wind_data$Date), 
                        tryFormats = c("%d-%m-%Y", "%d/%m/%Y"), 
                        origin="1970-01-01",tz="SAST")
head(my_base_date, n=4) 
```

## Pitfalls
- Format mistakes
- The format argument must match the *current* format of the date

```{r}
as.Date(as.character(wind_data$Date), 
        format = "%Y-%m-%d", tz="SAST", 
        origin="1970-01-01")
```
```{r, echo=TRUE, results='hide'}
as.Date(as.character(wind_data$Date), 
        format = "%Y/%m/%d", tz="SAST", 
        origin="1970-01-01")
```

```{r, echo=FALSE}
tt <- as.Date(as.character(wind_data$Date), 
        format = "%Y/%m/%d", tz="SAST", 
        origin="1970-01-01")
head(tt, n=3)
```

## Many format options...

```{r}
ymd("2020-01-31")

mdy("January 31st, 2020")

dmy("31-Jan-2020")

ymd(20200131)
```

## Date-times can also have lots of formats

```{r}
ymd_hms("2020-01-31 14:09:21")

mdy_hm("01/31/2020 06:09")

mdy_h("01/31/2020 06")
```

## Exercise 1

In `01-ts-exercise.R`.

Use the appropriate `lubridate` function to parse each of the following dates:

```
d1 <- "January 1, 2010"
d2 <- "2015-Mar-07"
d3 <- "06-Jun-2017"
d4 <- c("August 19 (2015)", "July 1 (2015)")
d5 <- "12/30/14" # Dec 30, 2014
```

## Time zones

- You can specify a time zone up-front
```{r}
ymd_hms("2020-01-31 14:09:21", tz="Africa/Johannesburg")
```
 
- Or add it later
```{r}
mytime <- ymd_hms("2020-01-31 14:09:21")
with_tz(mytime, tzone="Africa/Johannesburg")
```

## Once you have a date, time or <br /> date-time object
- You can extract elements from it, like the date

```{r}
womensday <- ymd_hms("2020-08-09-12-30-30", tz="GMT"); womensday
date(womensday); month(womensday)

# Is it a leap year?
leap_year(womensday)
```

## Once you have a date, time or <br /> date-time object
- Or elements of the time

```{r}
hour(womensday); minute(womensday); second(womensday)
```

## Once you have a date, time or <br /> date-time object
- Or elements of the calendar date

```{r}
mday(womensday); wday(womensday, label=TRUE); yday(womensday)
```

## You can also create datetimes <br /> from their components

```{r}
temp_data <- read.csv("data/temperature.csv")
head(temp_data, n=3)
```

## You can also create datetimes <br /> from their components

```{r}
temp_data %>% 
  select(Year, Month, Day, Hour, Minute, Second) %>% 
  mutate(DateTime = make_datetime(Year, Month, Day, 
                                  Hour, Minute, Second))
```

## Rounding dates
- Sometimes you might want to round a date

```{r}
floor_date(now(), unit="minutes")
ceiling_date(now(), unit="minutes")
```

## Exercise 2

In `02-ts-exercise.R`.

1. Open the `flights` dataset
2. Use `make_datetime()` to combine the year, month, day, hour and minute variables into a single variable containing the departure time.
3. Remove the variables you used to make up the departure variable.
4. Make a line plot of the average departure delay within each hour of the day.

## Time spans

- Durations: exact number of seconds between events
<br><br>
- Periods: track clock time between events in human units like weeks and months 
<br><br>
- Intervals: have a start and an end point

## Durations

- Durations: a time span in exact number of seconds
- Subtracting two dates can give confusing results
- Durations only measure time in seconds

How old is Kanye West?
```{r}
kw_age <- today() - ymd(19770608); kw_age
as.duration(kw_age)
```

## Durations

How old will Kanye West be next week?
```{r}
kw_age + dweeks(1)
as.duration(kw_age + dweeks(1))
```

## Periods

- Periods: accurately track clock time despite human time constructs like leap seconds, leap years and daylight savings 

```{r}
months(8) + days(12) + seconds(3)
```

```{r}
ymd("2020-01-01") + dyears(1)
#> [1] "2016-12-31 06:00:00 UTC"
ymd("2020-01-01") + years(1)
#> [1] "2017-01-01"
```

## Intervals
- Intervals: have a start and an end point

```{r}
# Create an interval in time
lunch <- paste(as.character(today()), "12-30-05")
lunchtime <- ymd_hms(lunch, tz="Africa/Johannesburg"); lunchtime
time_since_lunch <- lubridate::interval(lunchtime, now()); time_since_lunch

as.duration(time_since_lunch)
```

## Intervals
- Intervals are made up of two dates

```{r}
# Access the start of an interval
start <- int_start(time_since_lunch); start

# Access the end point of an interval
end <- int_end(time_since_lunch); end
```

## Exercise 3

In `03-ts-exercise.R`.

Write a function that given your birthday (as a date), returns how old you are in years.

## Explorating time series through visualisation

- Packages `tsibble`, `feasts`, `brolgar`, `dygraphs`
- Datasets: 
    - dolphin sightings at Cape Vidal, KZN
    - wage data from the `brolgar` package
    - purchase data from [Kaggle](https://www.kaggle.com/c/acquire-valued-shoppers-challenge)
- What is a time series? 
    - a list of numbers of interest, indexed by the times those numbers were recorded 
    - there is a special type of dataframe in `R` for time series, a `tsibble`

## Dolphin sightings
 
- Hourly number of dolphin sightings

```{r}
load("data/hourly_dolphin_counts.RData")
head(hourly_counts, n=5)
```

## Dolphin sightings
 
- Weekly number of dolphin sightings
- We only keep date to help us with plotting

```{r}
weekly_counts <- hourly_counts %>% 
    mutate(week=week(datetime)) %>% group_by(week, location) %>%
    summarise(sum_counts=sum(dolphin_counts),
              sum_effort=sum(effort),
              mean_sight=mean(sightability),
              date=mean(datetime)); head(weekly_counts, n=3)
```

## From dataframe to `tsibble`

```{r}
weekly_counts <- as_tsibble(x = weekly_counts,
                    key = location,
                    index = week,
                    regular = TRUE)
head(weekly_counts)
```

## Plotting a time series

```{r}
weekly_counts %>%
  ggplot(aes(x = date,
             y = sum_counts)) + geom_line() +
  ylab("Total weekly sightings") + xlab("Month") + 
  facet_grid(~ location)
```

## Seasonal decomposition

```{r}
sdecomp <- weekly_counts %>%
  model(STL(sum_counts ~ trend())) %>%
  components()

autoplot(sdecomp) + xlab("Week")
```

## Exercise 4

In `04-ts-exercise.R`.

Use the tools above to see if there are more purchases made on certain days of the week.

## Individuals through time with `brolgar`

- There is a dataset built into this package about wages

```{r}
head(wages)
```
 
## Individuals through time with `brolgar`

- As you can see it's a `tsibble` and it was created like this

```
wages <- as_tsibble(x = wages,
                    key = id,
                    index = xp,
                    regular = FALSE)
```

## Individuals through time with `brolgar`

- Really nice functionality for exploring datasets composed of time series from many individuals or groupings, e.g., 

  + individual humans or animals
  + individual locations
  + individual companies 
  + individual sports teams
  + ...

## Sample a handful of time series

```{r, tidy=FALSE}
wages %>%
  sample_n_keys(size = 5) %>%
  ggplot(aes(x = xp, y = ln_wages, group = id)) + 
  geom_line() + 
  labs(y="Log(wage)", x="Experience (number of years)")
```

## Sample a handful of time series

```{r, tidy=FALSE}
ggplot(wages,
       aes(x = xp, y = ln_wages, group = id)) +
       geom_line() + facet_sample(n_per_facet = 3, n_facets = 6) +
  labs(y="Log(wage)", x="Experience (number of years)")
```

## Exercise 5

In `05-ts-exercise.R`.

Repeat the analysis above for the `ChickWeight` dataset. Look for any interesting patterns in individual chicks.

## Interactive time series plots

```{r, tidy = FALSE}
weekly_counts %>% dplyr::select(location, week, sum_counts) %>% 
  pivot_wider(names_from = location, values_from = sum_counts) %>% 
  dygraph() %>% dyOptions(stackedGraph = FALSE) %>%
  dyRangeSelector(height = 20)
```

---

```{r, tidy = FALSE}
weekly_counts %>% dplyr::select(location, week, sum_counts) %>% 
  pivot_wider(names_from = location, values_from = sum_counts) %>% 
  dygraph() %>% dyOptions(stackedGraph = TRUE) %>%
  dyRangeSelector(height = 20)
```

## Exercise 6

In `06-ts-exercise.R`.

Use the `dygraphs` package to create an interactive plot of weekly total purchases for each brand A, B, C, D

## Reading

+ https://r4ds.had.co.nz/dates-and-times.html
+ https://cran.r-project.org/web/packages/feasts/vignettes/feasts.html
+ https://feasts.tidyverts.org/
+ https://www.njtierney.com/post/2019/08/13/introducing-brolgar/
