---
title: "Clustering"
author: Lecture XX
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.height=3.5,
                      tidy.opts=list(width.cutoff=60),
                      tidy=TRUE, 
                      warning = FALSE)
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(imager)
library(formatR)
```

```{r wrap-hook, echo=FALSE}
library(knitr)
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = knitr:::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})
knitr::opts_chunk$set(linewidth=60)
```

## Clustering

```{r}
ratings <- readRDS(file = "data/movieratings.Rds")
str(ratings)
```

```{r, echo = FALSE}
# reshape long to wide
ratings_wide <- ratings %>% 
  dplyr::select(userId, title, rating) %>% 
  pivot_wider(names_from = title, values_from = rating)
```

## Visualising the raw data

```{r, echo = FALSE, fig.height=6}
# visualize the raw data
par(mai = c(2,.5,.5,.5))
image(t(ratings_wide[,-1])[, (nrow(ratings_wide)-1):1], axes = F)
axis(1, at = 0:29/29, names(ratings_wide)[-1], las = 3, cex.axis = 0.7)
axis(2, at = 0:99/99, rownames(ratings_wide), cex.axis = 0.6)
```

## Heatmaps

```{r, echo = FALSE, fig.height=6}
heat <- heatmap(as.matrix((ratings_wide[,-1])), 
        scale = "none", margins = c(12, 5))
```

## Manual heatmaps with `image`

```{r,eval=FALSE}
# heatmap
heat <- heatmap(as.matrix((ratings_wide[,-1])), 
        scale = "none", margins = c(12, 5))
row_ordering <- heat$rowInd
col_ordering <- heat$colInd
# reshape long to wide
ratings_wide <- ratings %>% 
  dplyr::select(userId, title, rating) %>% 
  pivot_wider(names_from = title, values_from = rating)
# image wants a matrix of ratings
ratings_mat <- as.matrix(ratings_wide[,-1])
```

```{r,echo=FALSE}
# heatmap
row_ordering <- heat$rowInd
col_ordering <- heat$colInd
# reshape long to wide
ratings_wide <- ratings %>% 
  dplyr::select(userId, title, rating) %>% 
  pivot_wider(names_from = title, values_from = rating)
# image wants a matrix of ratings
ratings_mat <- as.matrix(ratings_wide[,-1])
```

--- 

```{r, fig.height=6}
image(t(ratings_mat[row_ordering, col_ordering]))
```

# How does it cluster?

## Hierarchical clustering with `hclust`

```{r, fig.height=4.5}
# single linkage
hcl_single <- dist(ratings_wide[,-1]) %>% hclust(method = "single")
plot(hcl_single)

```

---

```{r, fig.height=4.5}
# complete linkage
hcl_complete <- dist(ratings_wide[,-1]) %>% hclust(method = "complete")
plot(hcl_complete)
```

---
```{r, fig.height=4.5}
# centroid linkage
hcl_centroid <- dist(ratings_wide[,-1]) %>% hclust(method = "centroid")
plot(hcl_centroid)
```


## Where to cut?

```{r, echo=FALSE, fig.height=4.5}
plot(hcl_complete)
```

## Allocate each row to a cluster

```{r}
user_clusters <- cutree(hcl_complete, h = 9)
ratings_wide <- ratings_wide %>% mutate(user_clusters = user_clusters)
table(ratings_wide$user_clusters)
```

```{r, echo=FALSE}
# for later
user_clusters_df <- data.frame(userId = ratings_wide$userId, 
                               user_clus = user_clusters, row.names = NULL)
```

## Review clusters

```{r, echo = FALSE, fig.height=4.5}
# plot mean movie ratings within each cluster
cluster_means <- ratings_wide %>% 
  group_by(user_clusters) %>%
  summarise_at(vars(-starts_with("user")), mean, na.rm = TRUE) %>% 
  ungroup()

cluster_means %>% mutate(user_clusters = factor(user_clusters)) %>%
  pivot_longer(cols = -1, names_to = "title", values_to = "rating") %>%
  ggplot(aes(x = title, y = rating, colour = user_clusters, group = user_clusters)) + 
  geom_line() + coord_flip() + xlab(element_blank()) + ylab(element_blank())
```

---

```{r, echo = FALSE, fig.height=4.5}
cluster_means %>% mutate(user_clusters = factor(user_clusters)) %>%
  pivot_longer(cols = -1, names_to = "title", values_to = "rating") %>%
  ggplot(aes(x = title, y = rating, colour = user_clusters, group = user_clusters)) +
  facet_grid(. ~ user_clusters) +
  geom_line() + coord_flip() + xlab(element_blank()) + ylab(element_blank())
```

## Redo heatmap after grouping users into clusters

```{r, echo=FALSE, fig.height=4.5}
ratings_userclus <- ratings_wide %>% 
  dplyr::select(-userId) %>% group_by(user_clusters) %>%
  summarise_all(mean, na.rm = TRUE) 
heatmap(as.matrix((ratings_userclus[,-1])), scale = "none", margins = c(12, 5))

```

## ... or keep users separate but reorder

```{r, echo=FALSE, fig.height=6}
hcl_row_ordering <- hcl_complete$order
par(mfrow = c(1,2))
image(t(as.matrix(ratings_mat)), main = "Original Matrix")
image(t(ratings_mat[hcl_row_ordering, ncol(ratings_mat):1]), main = "Reordered rows")
```

# Clustering movies 

## Transpose and repeat...

```{r}
# transpose and leave out Id and cluster memb columns
ratings_wide_t <- t(ratings_wide[,-c(1,32)]) %>% as.data.frame()
# column names should be userId's
names(ratings_wide_t) <- ratings_wide$userId
# complete linkage again (or try others)
hcl_complete <- dist(ratings_wide_t) %>% hclust(method = "complete")
```

---

```{r, echo=FALSE,fig.height=6}
plot(hcl_complete)
```

## Allocate movies to clusters

```{r}
title_clusters <- cutree(hcl_complete, k = 6)
table(title_clusters)
```

```{r, echo=FALSE}
# for later
title_clusters_df <- data.frame(title = names(title_clusters), title_clus = title_clusters, row.names = NULL)
```

---

```{r, linewidth=60}
# movies in cluster 1
str_c(title_clusters_df[title_clusters_df$title_clus == 1, "title"], collapse = "; ")
# movies in cluster 2
str_c(title_clusters_df[title_clusters_df$title_clus == 2, "title"], collapse = "; ")
```

## Review clusters

```{r, echo=FALSE, fig.height=5}
ratings_wide_t <- ratings_wide_t %>% mutate(title_clusters = title_clusters)

ratings_titleclus <- ratings_wide_t %>% group_by(title_clusters) %>%
  summarise_all(mean, na.rm = TRUE) 

# plot mean movie ratings within each cluster
cluster_means <- ratings_wide_t %>% 
  group_by(title_clusters) %>%
  summarise_all(mean, na.rm = TRUE) %>% 
  ungroup()

cluster_means %>% mutate(title_clusters = factor(title_clusters)) %>%
  pivot_longer(cols = -1, names_to = "user", values_to = "rating") %>%
  ggplot(aes(x = user, y = rating, colour = title_clusters, group = title_clusters)) + 
  geom_line() + coord_flip() + xlab(element_blank()) + ylab(element_blank())

```

---

```{r, echo=FALSE, fig.height=6}
cluster_means %>% mutate(title_clusters = factor(title_clusters)) %>%
  pivot_longer(cols = -1, names_to = "user", values_to = "rating") %>%
  ggplot(aes(x = user, y = rating, colour = title_clusters, group = title_clusters)) + 
  geom_line() + coord_flip() + xlab(element_blank()) + ylab(element_blank()) +
  facet_grid(. ~ title_clusters)
```

## Heatmap with movie clusters

```{r, echo=FALSE, fig.height=4.5}
heatmap(as.matrix((ratings_titleclus[,-1])), scale = "none", margins = c(12, 5))
```

## ... or just reordering movies

```{r, echo=FALSE, fig.height=6}
hcl_col_ordering <- hcl_complete$order
par(mfrow = c(1,2))
image(t(as.matrix(ratings_mat)), main = "Original Matrix")
image(t(ratings_mat[nrow(ratings_mat):1, col_ordering]), main = "Reordered columns")
```

# Putting it all together

---

```{r, echo=FALSE}
par(mfrow = c(1,4))
image(t(as.matrix(ratings_mat)), main = "Original Matrix")
image(t(ratings_mat[hcl_row_ordering, ncol(ratings_mat):1]), main = "Reordered rows")
image(t(ratings_mat[nrow(ratings_mat):1, hcl_col_ordering]), main = "Reordered columns")
image(t(ratings_mat[hcl_row_ordering, hcl_col_ordering]), main = "Reordered both")
```

## World's smallest heatmap

```{r, echo=FALSE,fig.height=4.5}
# merge in cluster memberships for users and movies
ratings <- ratings %>% 
  left_join(user_clusters_df, by = "userId") %>% 
  left_join(title_clusters_df, by = "title")

# summarize
ratings_bothclus <- ratings %>% 
  group_by(title_clus, user_clus) %>%
  summarize(rating = mean(rating, na.rm = TRUE)) %>% 
  ungroup() %>% 
  pivot_wider(names_from = title_clus, values_from = rating)

heatmap(as.matrix((ratings_bothclus[,-1])), scale = "none", margins = c(5, 5))
```

# k-means clustering

---

```{r}
df <- ratings_wide %>% dplyr::select_at(vars(-starts_with("user")))
kmeansObj <- kmeans(df, centers = 6)
```
```{r}
names(kmeansObj)
```

--- 

```{r}
kmeansObj$cluster
```

```{r}
kmeansObj$centers[,1:3]
```

## Plotting cluster means

```{r}
# tidy for ggplot
kmm <- kmeansObj$centers %>% as.data.frame() %>% 
  mutate(user_clus_km = paste0("clust",1:6)) %>%
  pivot_longer(-user_clus_km, names_to = "title", values_to = "rating") 
head(kmm)
```

---

```{r, echo=FALSE}
kmm %>%
  ggplot(aes(x = title, y = rating, colour = user_clus_km, group = user_clus_km)) + 
  geom_line() + coord_flip() + xlab(element_blank()) + ylab(element_blank())
```

## similarity between `hclust` and `kmeans`

```{r}
table(user_clusters, kmeansObj$cluster)
```

# Dimension reduction

## Dimension reduction = clustering for variables

```{r}
# don't need to reorder rows and cols, but will make visualizing PCA easier
ratings_scaled_ordered <- scale(ratings_mat)[hcl_row_ordering, hcl_col_ordering]

# do the svd
svd1 <- svd(ratings_scaled_ordered)

# rename svd outputs
u <- svd1$u
v <- svd1$v
d <- diag(svd1$d)

# approximate original data with outer product of first singular vector
approx1 <- u[,1] %*% matrix(d[1,1],nrow=1) %*% t(v[,1])
```

---

```{r,echo=FALSE,fig.height=6}
par(mfrow = c(1, 2))
image(t(ratings_scaled_ordered[100:1,]), main = "Original Matrix")
image(t(approx1)[, nrow(approx1):1], main = "Approximation (1 SV)")
```

---

```{r}
approx2 <- u[,1:2] %*% d[1:2,1:2] %*% t(v[,1:2])
approx5 <- u[,1:5] %*% d[1:5,1:5] %*% t(v[,1:5])
approx30 <- u %*% d %*% t(v)
```

---

```{r,echo=FALSE,fig.height=6}
par(mfrow = c(1, 5))
image(t(ratings_scaled_ordered[100:1,]), main = "Original Matrix")
image(t(approx1)[, nrow(approx1):1], main = "Approximation (1 SV)")
image(t(approx2)[, nrow(approx2):1], main = "Approximation (2 SV)")
image(t(approx5)[, nrow(approx5):1], main = "Approximation (5 SV)")
image(t(approx30)[, nrow(approx30):1], main = "Approximation (30 SV)")
```
 
## How many vectors are needed?

```{r, echo=FALSE, fig.height=6}
par(mfrow=c(1,3))
plot(svd1$d, xlab = "Column", ylab = "Singular value", pch = 19)
plot(svd1$d / sum(svd1$d), xlab = "Column", ylab = "% variance explained", pch = 19)
plot(cumsum(svd1$d) / sum(svd1$d), xlab = "Column", ylab = "Cumulative % variance explained", pch = 19)

```

## SVD vs PCA

PCs are equal to the right SVs if you scale the data to have mean 0 and sd 1

---

```{r, fig.height=4}
ratings_ordered <- ratings_mat[hcl_row_ordering, hcl_col_ordering]
pca1 <- prcomp(ratings_ordered, scale = TRUE)
plot(pca1$rotation[, 1], svd1$v[, 1])
```

---

```{r, echo=FALSE, fig.height=5}
par(mfrow = c(1,2))
plot(svd1$d, xlab = "Column", pch = 19)
plot(pca1$sdev, xlab = "Column", pch = 19)
```

# Image compression with PCA
---

```{r, echo=FALSE}
mona <- load.image("data/lowres_mona.png")
plot(mona)
```

---

```{r}
# transform into matrix form
mona_mat <- as.data.frame(mona) %>% 
  pivot_wider(names_from = y, values_from = value) %>%
  dplyr::select(-x) %>%
  as.matrix()
image(mona_mat[, nrow(mona_mat):1], col = gray.colors(100))
```

---

```{r}
# svd
svd1 <- svd(mona_mat)
u <- svd1$u; v <- svd1$v; d <- diag(svd1$d)

# number of singular values
nsv <- 1
# approximate original data with outer product of first N singular vectors
approx <- u[,1:nsv] %*% matrix(d[1:nsv,1:nsv],nrow=nsv) %*% t(v[,1:nsv])
```

## With 1 SV

```{r, echo=FALSE,fig.height=5}
image(approx[, nrow(approx):1], col = gray.colors(30))
```

## With 2 SV

```{r, echo=FALSE,fig.height=5}
nsv <- 2
approx <- u[,1:nsv] %*% matrix(d[1:nsv,1:nsv],nrow=nsv) %*% t(v[,1:nsv])
image(approx[, nrow(approx):1], col = gray.colors(30))
```

## With 5 SV

```{r, echo=FALSE,fig.height=5}
nsv <- 5
approx <- u[,1:nsv] %*% matrix(d[1:nsv,1:nsv],nrow=nsv) %*% t(v[,1:nsv])
image(approx[, nrow(approx):1], col = gray.colors(30))
```

## With 20 SV

```{r, echo=FALSE,fig.height=5}
nsv <- 20
approx <- u[,1:nsv] %*% matrix(d[1:nsv,1:nsv],nrow=nsv) %*% t(v[,1:nsv])
image(approx[, nrow(approx):1], col = gray.colors(30))
```

## With 50 SV

```{r, echo=FALSE,fig.height=5}
nsv <- 50
approx <- u[,1:nsv] %*% matrix(d[1:nsv,1:nsv],nrow=nsv) %*% t(v[,1:nsv])
image(approx[, nrow(approx):1], col = gray.colors(30))
```

## How many SVs is enough?

```{r, echo=FALSE,fig.height=5}
plot(svd1$d / sum(svd1$d), xlab = "Column", ylab = "% variance explained", pch = 19)
```

