---
title: "Exploring spatial data"
author: Lecture XX
output: ioslides_presentation
---


```{r, include = FALSE}
knitr::opts_chunk$set(fig.height=3.5) 
library(dplyr)
library(ggplot2)
library(sf)
library(leaflet)
library(stringr)
library(tmap)
library(mgcv)
## get occupancy results
occupancies <- st_read("data/Mongolia_SL_Occ.shp")
occupancies %>% st_crs()
occupancies <- occupancies %>% st_set_crs("+proj=utm +zone=48 +datum=WGS84 +units=m +no_defs")

###### generate camera trap locations from a lat-long file 

## get camera locations
cams <- read.csv("data/camera-locations.csv")
cams <- st_as_sf(cams, coords = c("Longitude", "Latitude"), crs = 4326) %>% 
  st_transform(crs = "+proj=utm +zone=48 +datum=WGS84 +units=m +no_defs")
```

# Introducing spatial data

## Spatial data types

- Vector data
  + represents space using points, lines and polygons ("features")
  + each of these can have additional data ("attributes")
  + R package `sf`
<br><br>
- Raster data
  + represents space using cells of equal size
  + each cell can have additional data ("layers")
  + R package `raster`

## Vector data

## Raster data

# Simple features

## What is a feature?

- A **feature** is a thing... a city, a house, a tree
  + has a **geometry** describing where it is located
  + has **attributes** describing its other properties
<br><br>
- A feature may be made up of other features
  + pixel vs. image

## What are simple features?

- open standard for spatial data types
<br><br>
- points, lines and polygons = "geometries"
<br><br>
- geometries are hierarchical
  + set of points is a line
  + set of lines forms a polygon
  + several polygons is a multipolygon
<br><br>
- 17 geometries, of which we look at 7
<br><br>
- ["data frames with a spatial extension"](https://geocompr.robinlovelace.net/spatial-class.html#sfc)

## `sf` geometry types

- A point: `st_point()`
- A linestring: `st_linestring()`
- A polygon: `st_polygon()`
- A multipoint: `st_multipoint()`
- A multilinestring: `st_multilinestring()`
- A multipolygon: `st_multipolygon()`
- A geometry collection: `st_geometrycollection()`
<br><br>
- Can be created from vectors, matrices, or lists

## Features made of single geometries

```{r}
# POINT
(pt1 <- st_point(c(5, 2))) 
```

```{r, echo = FALSE}
plot(pt1)
title("POINT", line = 0.1)
box()
```

---

```{r}
# LINESTRING
(ls1 <- st_linestring(rbind(c(5, 2), c(1, 3), c(3, 4))))           
```

```{r, echo = FALSE}
plot(ls1)
title("LINESTRING", line = 0.1)
box()
```

---

```{r}
# POLYGON
pl <- list(rbind(c(5, 2), c(1, 3), c(3, 4), c(5, 2)))
(poly1 <- st_polygon(pl))  
```

```{r, echo = FALSE}
plot(poly1)
title("POLYGON", line = 0.1)
box()
```

## Features made of multiple geometries 

```{r}
# MULTIPOINT
mpmat <- rbind(c(5, 2), c(1, 3), c(3, 4))
(mp1 <- st_multipoint(mpmat))
```

```{r, echo = FALSE}
plot(mp1)
title("MULTIPOINT", line = 0.1)
box()
```

---

```{r}
# MULTILINESTRING
mll <- list(rbind(c(1, 5), c(4, 4), c(4, 1), c(2, 2)), 
           rbind(c(1, 2), c(2, 4)))
(mls1 <- st_multilinestring((mll)))
```

```{r, echo = FALSE}
plot(mls1)
title("MULTILINESTRING", line = 0.1)
box()
```

---

```{r}
# MULTIPOLYGON
mpyl <- list(list(rbind(c(1, 5), c(2, 2), c(4, 1), c(1, 5))),
            list(rbind(c(0, 2), c(1, 2), c(1, 3), c(0, 3), c(0, 2))))
(mpoly1 <- st_multipolygon(mpyl))
```

```{r, echo = FALSE}
plot(mpoly1)
title("MULTIPOLYGON", line = 0.1)
box()
```

## Combination of geometries

```{r}
# GEOMETRYCOLLECTION
gcl <- list(ls1, pt1, mpoly1)
(geomcol <- st_geometrycollection(gcl))
```

```{r, echo = FALSE}
plot(geomcol)
title("GEOMETRYCOLLECTION", line = 0.1)
box()
```

# Building up spatial data frames in R

## From `sfg` to `sfc` 

- Objects we've made so far have class `sfg` 

```{r}
class(pt1)
class(mp1)
```

- A **simple feature geometry column** (sfc) is a list of sfg objects
- Important because usually we have more than one feature

## Combine simple features with `st_sfc()` 

```{r}
# sfc POINT
pt1 <- st_point(c(5, 2))
pt2 <- st_point(c(1, 3))
(points_sfc <- st_sfc(pt1, pt2))
```

---

```{r}
class(points_sfc)
is.list(points_sfc) # is this a list?
is.list(mp1)        # is this (multipoint) a list?
st_geometry_type(points_sfc)
```

## From `sfc` to `sf` 

- `sfg` and `sfc` only tell you about locations
<br><br>
- to these we add non-geographic attributes associated with each feature
<br><br>
- an `sf` object is a data frame combines these two things:
  + attributes are stored in a regular data.frame
  + location information (co-ordinates) is stored in one (simple feature geometry) column (`sfc`) of this data frame
<br><br>
- Simple features = data frames with spatial attributes stored in a list-column

---

```{r}
lnd_point = st_point(c(0.1, 51.5))                 # sfg object
lnd_geom = st_sfc(lnd_point, crs = 4326)           # sfc object
lnd_attrib = data.frame(                           # data.frame object
  name = "London",
  temperature = 25,
  date = as.Date("2017-06-21")
  )
lnd_sf = st_sf(lnd_attrib, geometry = lnd_geom)    # sf object
lnd_sf
```

# Coordinate reference systems

## Reprojecting 

# Reading in spatial data

## From a shape file

```{r}
occupancies <- st_read("data/Mongolia_SL_Occ.shp")
occupancies %>% st_crs()     # check the CRS
```

```{r}
occupancies <- occupancies %>% st_set_crs("+proj=utm +zone=48 +datum=WGS84 +units=m +no_defs")
occupancies %>% st_crs()     
```

## From a spreadsheet or non-spatial data frame

```{r}
# g
cams <- read.csv("data/camera-locations.csv")
cams <- st_as_sf(cams, coords = c("Longitude", "Latitude"), crs = 4326)
cams %>% st_crs()
```

```{r}
# reprojecting
cams <- cams %>% 
  st_transform(crs = "+proj=utm +zone=48 +datum=WGS84 +units=m +no_defs")
```



<!-- # Static maps -->

<!-- ## tmap -->

<!-- - similar syntax to ggplot2, tailored to maps -->
<!-- - shape is a spatial object with class `sf`, `sp` or `raster`.  -->
<!-- ```{r, fig.height = 2.5} -->
<!-- map0 <- tm_shape(occupancies) + tm_fill()  -->
<!-- map1 <- tm_shape(occupancies) + tm_borders()  -->
<!-- tmap_arrange(map0, map1, nrow = 1) -->
<!-- ``` -->

<!-- --- -->

<!-- ```{r} -->
<!-- map2 <- tm_shape(occupancies) + tm_polygons()  -->
<!-- map3 <- tm_shape(occupancies) + tm_polygons("Occ_1200")  -->
<!-- tmap_arrange(map2, map3, nrow = 1) -->
<!-- ``` -->

<!-- --- -->

<!-- - Multiple layers per shape  -->
<!-- ```{r} -->
<!-- tm_shape(occupancies) +  -->
<!--   tm_polygons("Occ_1200") +  -->
<!--   tm_text("Team", size = 0.4) -->
<!-- ``` -->

<!-- --- -->

<!-- - Multiple shapes -->
<!-- ```{r, fig.cap="Color settings are especially important for maps. Here use of the same palette obscures detail."} -->
<!-- tm_shape(occupancies) + tm_polygons("Occ_1200") + -->
<!--   tm_shape(cams) + tm_dots("Elevation", size = 0.5) -->
<!-- ``` -->

<!-- --- -->

<!-- - Sequential palettes are good for values that have natural ordering from low to high -->
<!-- ```{r, echo=TRUE} -->
<!-- tm_shape(occupancies) + tm_polygons("Occ_1200") +  -->
<!--   tm_shape(cams) + tm_dots("Elevation", size = 0.5, palette = "BuGn") -->
<!-- ``` -->

<!-- --- -->

<!-- - Categorical palettes are best for data with no natural ordering -->
<!-- ```{r, echo=TRUE} -->
<!-- tm_shape(occupancies) + tm_polygons("Occ_1200") +  -->
<!--   tm_shape(cams) + tm_dots("Mountain", size = 0.5, palette = "Set3") -->
<!-- ``` -->

<!-- --- -->

<!-- - Diverging palettes emphasise differences from some central reference point -->
<!-- ```{r, echo=TRUE} -->
<!-- tm_shape(occupancies) + tm_polygons("Occ_1200") +  -->
<!--   tm_shape(cams) + tm_dots("Elevation", size = 0.5, palette = "PRGn") -->
<!-- ``` -->

<!-- --- -->

<!-- ```{r, echo = FALSE, fig.height = 7} -->
<!-- library(RColorBrewer) -->
<!-- display.brewer.all() -->
<!-- ``` -->

<!-- ---  -->

<!-- - faceting -->
<!-- ```{r} -->
<!-- tm_shape(occupancies) + -->
<!--     tm_polygons(c("Occ_1200", "Team")) + -->
<!--     tm_facets(sync = TRUE, ncol = 2) -->
<!-- ``` -->

<!-- ## ggplot -->

<!-- - `geom_sf` plots `sf` objects -->
<!-- ```{r} -->
<!-- ggplot() + geom_sf(data = occupancies)   -->
<!-- ``` -->

<!-- --- -->

<!-- - `geom_sf` plots `sf` objects -->
<!-- ```{r} -->
<!-- ggplot() + geom_sf(data = occupancies, aes(fill = Occ_1200))  -->
<!-- ``` -->

<!-- --- -->

<!-- - Can plot multiple `sf` objects -->
<!-- ```{r} -->
<!-- ggplot() + geom_sf(data = occupancies, aes(fill = Occ_1200)) + -->
<!--   geom_sf(data = cams, colour = "red")  -->
<!-- ``` -->

<!-- --- -->

<!-- - `scale_XXX_brewer` for discrete scales, `scale_XXX_distiller` for continuous scales -->
<!-- ```{r} -->
<!-- ggplot() + geom_sf(data = occupancies, aes(fill = Occ_1200)) + -->
<!--   geom_sf(data = cams, aes(colour = Elevation)) + -->
<!--   scale_colour_distiller(palette = "PRGn") -->
<!-- ``` -->

<!-- # Interactive maps -->

<!-- ## package `leaflet` -->

<!-- - provides an interface to the Leaflet JavaScript library -->
<!-- - `leaflet()` makes a map object that can be piped to other leaflet functions -->
<!-- ```{r} -->
<!-- leaflet() %>% addTiles() -->
<!-- ``` -->

<!-- --- -->

<!-- - Set a starting origin and zoom level  -->
<!-- ```{r} -->
<!-- leaflet() %>% addTiles() %>% -->
<!--   setView(lng = 94.1, lat = 47.4, zoom = 7) -->
<!-- ``` -->

<!-- --- -->

<!-- - Set a starting origin and zoom level  -->
<!-- ```{r} -->
<!-- leaflet() %>% addTiles() %>% -->
<!--   setView(lng = 94.1, lat = 47.4, zoom = 4) -->
<!-- ``` -->

<!-- --- -->

<!-- - Various base map options with `addProviderTiles()` -->
<!-- ```{r} -->
<!-- leaflet() %>% addProviderTiles("Esri.WorldStreetMap") %>% -->
<!--   setView(lng = 94.1, lat = 47.4, zoom = 4) -->
<!-- ``` -->

<!-- --- -->

<!-- - Various base map options with `addProviderTiles()` -->
<!-- ```{r} -->
<!-- leaflet() %>% addProviderTiles("Esri.WorldImagery") %>% -->
<!--   setView(lng = 94.1, lat = 47.4, zoom = 4) -->
<!-- ``` -->

<!-- --- -->

<!-- - `sf` objects can be passed as additional map layers -->
<!-- - must be in lat/long format -->
<!-- ```{r} -->
<!-- ### <b>  -->
<!-- leaflet(data = st_transform(occupancies, crs = 4326)) %>% -->
<!--   ### </b>  -->
<!--   addProviderTiles("Esri.WorldImagery") %>% -->
<!--   addPolygons()  -->
<!-- ``` -->

<!-- --- -->

<!-- - Multiple map layers  -->
<!-- ```{r} -->
<!-- leaflet(data = st_transform(occupancies, crs = 4326)) %>% -->
<!--   addProviderTiles("Esri.WorldImagery") %>% -->
<!--   addPolygons() %>% -->
<!--   ### <b>  -->
<!--   addMarkers(data = st_transform(cams, crs = 4326)) -->
<!--   ### </b> -->
<!-- ``` -->

<!-- --- -->

<!-- ```{r} -->
<!-- ### <b>  -->
<!-- pal <- colorNumeric("YlOrRd", domain = occupancies$Occ_1200) -->
<!-- ### </b>  -->
<!-- leaflet(st_transform(occupancies, crs = 4326)) %>%  -->
<!--   addProviderTiles("Esri.WorldImagery") %>% -->
<!--   addMarkers(data = st_transform(cams, crs = 4326)) %>% -->
<!--   ### <b>  -->
<!--   addPolygons(color = ~pal(Occ_1200), fillOpacity = 0.4) %>% -->
<!--   addLegend(pal = pal, values = ~Occ_1200) %>% -->
<!--   addMiniMap(position = "bottomleft") -->
<!-- ### </b> -->
<!-- ``` -->

<!-- # Further reading -->

<!-- ---  -->

<!-- - [Chapter 2](https://geocompr.robinlovelace.net/spatial-class.html) and [Chapter 8](https://geocompr.robinlovelace.net/adv-map.html) of Geocomputation with R -->
<!-- - [sf vignette](https://cran.r-project.org/web/packages/sf/vignettes/sf1.html#reading_and_writing) -->
<!-- - [tmap vignette](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html) -->
